---
title: "04 - figuras"
author: "Elio Campitelli"
output: 
   powerpoint_presentation:
    # ioslides_presentation:
        fig_height: 5
        fig_width: 12
        #reference_doc: ppttemplate2.pptx
---

```{r setup, include=FALSE}
# Notification
start.time <- unclass(Sys.time())
min.time <- 10
knit_doc <- knitr::knit_hooks$get("document")
knitr::knit_hooks$set(document = function(x) {
   took <- unclass(Sys.time()) - start.time
   if (unclass(Sys.time()) - start.time >= min.time) {
      notify("Done knitting!", 
             paste0("Took ", round(took), " seconds"),
             time = 5)
   }  
   knit_doc(x)
})

knitr::opts_chunk$set(echo = FALSE,
                      cache = TRUE, cache.lazy = TRUE,
                      warning = FALSE)

library(data.table)
library(ggplot2)
library(dplyr)
library(metR) 
library(circular)
library(hrbrthemes)
library(extrafont)
library(patchwork)
library(officer)

knit_error <-  knitr::knit_hooks$get("error")
knitr::knit_hooks$set(error = function(x, options) {
   notify("Error knitting!", time = 5)
   knit_error(x, options)
})


source("scripts/helperfun.R")
# Plot thingys

data.world <- BuildMap(res = 1, smooth = 1)
map.world <- geom_map2(data.world)
map.SH <- geom_map2(data.world[lat %b% c(-90, 20)], color = "gray20")


pres <- ReadNetCDF("DATA/srfp.mon.mean.nc")
pres.mean <- pres[, .(pres = median(pres)), by = lat]
pres.mean <- rbind(data.table(lat = 0.0, pres = Inf), 
                   pres.mean, 
                   data.table(lat = -90.0, pres = Inf))
surface <- geom_polygon(data = pres.mean, aes(y = pres), fill = "white", 
                        alpha = 1, color = "gray30", size = 0.5)
pres <- pres[, .(pres = mean(pres)), by = .(lon, lat)]

# From https://github.com/hrbrmstr/hrbrthemes/issues/18
d <- read.csv(extrafont:::fonttable_file(), stringsAsFactors = FALSE)
d[grepl("Light", d$FontName),]$FamilyName <- font_rc_light
write.csv(d,extrafont:::fonttable_file(), row.names = FALSE)
extrafont::loadfonts()

theme_elio <- theme_minimal(base_size = 11) +
   theme(
      text = element_text(family = font_rc),
      legend.position = "bottom", legend.box = "vertical",
      panel.spacing.y = unit(5, "mm"),
      panel.spacing.x = unit(5, "mm"),
      legend.spacing = unit(2, "mm"),
      plot.margin = grid::unit(rep(3, 4), "mm"),
      legend.title = element_blank(),
      legend.box.spacing = unit(3, "mm"),
      legend.margin = margin(t = -5),
      panel.grid = element_line(color = "gray50", size = 0.2, linetype = 3),
      panel.ontop = TRUE)
theme_set(theme_elio)
update_geom_defaults(metR:::GeomTextContour, list(family = font_rc))

update_geom_defaults("contour2", list(color = "black"))
update_stat_defaults("contour2", aes(linetype = factor(-sign(..level..))))

options(ggplot2.continuous.fill = "viridis", 
        ggplot2.continuous.color = "viridis")

coord_quickmap <- function(..., ylim = c(-90, -15)) {
   ggplot2::coord_quickmap(ylim = ylim, ...) 
} 

geom_contour_ <- function(..., gap = 0, rotate = FALSE) {
   if (gap != 0) {
      list(geom_contour2(..., gap = gap),
           geom_text_contour(..., rotate = rotate))
   } else {
      geom_contour2(...)
   }
}

# For vertical cross-sections
coord_latlev <- function(ratio = 20, ...) coord_fixed(ratio = ratio, ...)
coord_lonlev <- function(ratio = 20*4, ...) coord_fixed(ratio = ratio, ...)

lev.breaks <- c(10, 30, 100, 200, 500, 1000)

season <- function(...) {
   metR::season(..., lang = "en")
}


## Options for generating ppt output
ppt <- FALSE
pptfile <- "figuras.pptx"
ppttemplate <- "ppttemplate.pptx"
# Initial ppt
if (ppt == TRUE){
   my_pres <- read_pptx(ppttemplate)
   print(my_pres, pptfile)   
}

# Add a plot to ppt 
knit_plot <- knitr::knit_hooks$get("plot")

# knitr::knit_hooks$set(plot = function(x, options) {
#    if (ppt == TRUE) {
#       # print(options)
#       if (inherits(last_plot(), "gg")) {
#          read_pptx(pptfile)  %>%
#             add_slide(layout = "figure", master = "Office Theme") %>%
#             ph_with_gg(last_plot(), type = "pic") %>%
#             ph_with_text(options$fig.cap, type = "body") %>%
#             print(pptfile)
#          set_last_plot(NULL)   # remove last_plot()
#       } else {
#          read_pptx(pptfile ) %>%
#             add_slide(layout = "figure", master = "Office Theme") %>%
#             rvg::ph_with_vg(code = eval(parse(text = options$code)), 
#                             type = "pic") %>%
#             ph_with_text(options$fig.cap, type = "body") %>% 
#             print(pptfile)
#       }
#    }
#    knit_plot(x, options)
# })

# Turn off cache for figures
knitr::opts_hooks$set(fig.cap = function(options) {
   if (ppt == TRUE) options$cache <- FALSE
   options
})

```

# Persistencia y eventos.

```{r read-ncep}
ncep.f <- memoise::memoise(function(lat = -90:40, 
                                    lon = 0:360,
                                    time = lubridate::as_datetime(c("1979-12-01", "2015-12-01")),
                                    level = 10:1000,
                                    vars = "gh"){
   subset <- list(lat = lat, lon = lon, level = level,
                  time = time)
   n <- ReadNetCDF("DATA/hgt.mon.mean.nc", vars = c(gh = "hgt"),
                   subset = subset) %>% 
      setnames(., c("level"), c("lev"))
   if ("u" %in% vars) {
      n[, u := ReadNetCDF("DATA/uwnd.mon.mean.nc", out = "vector",
                          subset = subset)[[1]]] 
   }
   if ("v" %in% vars) {
      n[, v := ReadNetCDF("DATA/vwnd.mon.mean.nc", out = "vector",
                          subset = subset)[[1]]]
   }
   n[, time := as.Date(time[1]), by = time]
   return(n)
}, cache = memoise::cache_filesystem(".rcache"))
ncep <- ncep.f(level = 200)
```


```{r calc-qs.eof}
# ncep <- ncep.f(level = 200)
region <- expand.grid(lat = c(-65, -40),
                      lev = c(100, 700))
lats.eof <- c(-80, -30)
qs.eof <- copy(ncep[lev == 200 & lat %between% lats.eof]) %>%
   .[, gh.minus := FilterWave(gh, k = 0:-1), by = .(lat, time, lev)] %>% 
   .[, gh := gh.minus*sqrt(cos(lat*pi/180))] %>%
   .[, gh := Detrend(gh, time),
     by = .(lon, lat)] %>%
   EOF(data = ., gh ~ time | lon + lat, n = 1:10, B = 5000)

qs.eof$right[PC == "PC1", gh := -gh]
qs.eof$left[PC == "PC1", gh := -gh]
```

```{r}
ggplot(cut(qs.eof, 1:2)$right, aes(lon, lat)) +
   geom_contour_fill(aes(z = gh), breaks = AnchorBreaks(0, 0.01, 0)) +
   map.SH +
   scale_s_map() + 
   scale_fill_divergent(guide= guide_colorstrip_bottom()) +
   coord_quickmap() +
   facet_wrap(~PC)
```


```{r calc-events}
events <- cut(qs.eof, 1:2)$left %>%
   .[, gh.norm := (gh - mean(gh))/sd(gh), by = PC] %>% 
   .[, event := cut(gh.norm, c(-Inf, -0.5, 0.5, Inf), 
                    c("negative", "neutral", "positive"),
                    ordered_result = TRUE)] %>% 
   # cut.eof(1:2) %>% 
   .[, .(time, PC, event)] 

events.number <- events %>% 
   .[, .N, by = .(month(time), event, PC)] %>% 
   tidyr::complete(month, PC, event, fill = list(N = 0)) %>% 
   setDT() %>% 
   .[event != "neutral"] %>% 
   .[as.numeric(PC) %in% 1:2]

g <- events %>%
   .[ncep, on = c("time"), allow.cartesian = TRUE] %>%
   .[, .(gh = mean(gh)), by = .(lon, lat, month(time), event, PC)] %>% 
   dcast(lon + lat + month + PC ~ event, value.var = "gh") %>% 
   .[, `:=`(positive = positive - neutral, 
            negative = negative - neutral,
            neutral = NULL)]  %>% 
   melt(id.vars = c("lon", "lat", "month", "PC"), 
        variable.name = "event") %>% 
   .[events.number, on = c("month", "PC", "event")] %>% 
   # .[PC == "PC1"] %>%
   # .[month == 1] %>%
   ggplot(aes(lon, lat)) +
   geom_contour_fill(aes(z = value), breaks = AnchorBreaks(0, 20, 0)) +
   map.SH +
   stat_subset(aes(subset = lon == 60 & lat == -20, 
                   label = paste0("N = ", N)),
               geom = "label", size = 2) +
   scale_s_map() +
   coord_quickmap() +
   scale_fill_divergent(breaks = AnchorBreaks(0, 20, 0),
                        guide = guide_colorstrip_bottom())
```

```{r number-of-events, fig.cap = "Number of positive and negative events (+- 0.5 sigma) for each PC and each month."}
ggplot(events.number, aes(month, N)) +
   # geom_area(aes(ymax = N, fill = event), 
   #           position = "dodge", alpha = 0.5) +
   geom_line(aes(color = event)) +
   scale_color_brewer(palette = "Set1") +
   scale_x_continuous("Month", 1:12, labels = month.abb) +
   scale_y_continuous("# of events", limits = c(0, NA)) +
   facet_wrap(~PC, ncol = 1)
```

::: notes

Se ve clarmaente un ciclo anual (que es obvio por las divisones anteriores). Lo importante es ver que hay muy pocos eventos en algunas categorías, como positivos en PC1 verano o negativos en PC2 invierno. 

:::

```{r comp-events-1, fig.cap = "Mean gh anomaly (positive - neutral; negative - neutral) for positive and negatives events for each PC and each month."}
g + facet_grid_paginate(month ~ PC + event, ncol = 4, 
                        nrow = 6, page = 1,
                        labeller = labeller(month = month.abb))
```

```{r comp-events-2, fig.cap = "Mean gh anomaly (positive - neutral; negative - neutral) for positive and negatives events for each PC and each month. (continued)"}
g + facet_grid_paginate(month ~ PC + event, ncol = 4, 
                        nrow = 6, page = 2,
                        labeller = labeller(month = month.abb))
```

```{r}
oni <- fread("DATA/ONI_v5.csv") %>% 
   set_colnames(c("year", 1:12)) %>% 
   melt(id.vars = "year", variable.name = "month", value.name = "ENSO") %>% 
   .[, time := as.Date(paste0(year, "-", month, "-01"))] %>% 
   .[, c("year", "month") := NULL]

# ENSO events: oni > 0.5
oni[, enso.event := cut(ENSO, c(-Inf, -0.5, 0.5, Inf), c("negative", "neutral", "positive"))]

oni[events, on = "time"] %>% 
   .[complete.cases(.)]  %>%
   {
         chis <<- .[, broom::tidy(chisq.test(table(enso.event, event))), by = PC] 
         .
   }  %>% 
   .[, .N, by = .(enso.event, event, PC)] %>% 
   ggplot(aes(enso.event, event)) +
   geom_raster(aes(fill= N)) +
   geom_label(aes(label = N)) +
   scale_y_discrete("PC event") + 
   scale_x_discrete("ENSO event") +
   facet_wrap(~PC)
```

```{r}
knitr::kable(chis)
```



```{r}
region <- expand.grid(lat = c(-65, -40),
                      lev = c(100, 700))
lats.eof <- c(-80, -30)
i <- 1i
qs.ceof <- copy(ncep[lev == 200 & lat %between% lats.eof]) %>%
   .[, gh.minus := FilterWave(gh, k = 0:-1), by = .(lat, time, lev)] %>% 
   .[, gh := gh.minus*sqrt(cos(lat*pi/180))] %>%
   .[, gh := Detrend(gh, time),
     by = .(lon, lat)] %>%
   .[, c("R", "I") := ReIm(spectral::analyticFunction(gh)), 
     by = .(lat, time)] %>% 
   .[, gh := R + I*i] %>% 
   EOF(data = ., gh ~ time | lon + lat, n = 1:10, B = 5000, suffix = "cPC")
```


```{r}
events <- cut(qs.ceof, 1)$left %>%
   .[, gh := abs(gh)] %>% 
   .[, gh.norm := (gh - mean(gh))/sd(gh), by = cPC] %>% 
   .[, event := cut(gh.norm, c(-Inf, -0.5, 0.5, Inf), 
                    c("negative", "neutral", "positive" ),
                    ordered_result = TRUE)] %>%
   .[, .(time, cPC, event)] 

events.number <- events %>% 
   .[, .N, by = .(month(time), event, cPC)] %>% 
   tidyr::complete(month, cPC, event, fill = list(N = 0)) %>% 
   setDT() %>% 
   .[event != "neutral"] %>% 
   .[as.numeric(cPC) %in% 1]
```


```{r number-of-events-c, fig.cap = "Number of positive and negative events (+- 0.5 sigma) for each PC and each month."}
ggplot(events.number, aes(month, N)) +
   # geom_area(aes(ymax = N, fill = event), 
   #           position = "dodge", alpha = 0.5) +
   geom_line(aes(color = event)) +
   scale_color_brewer(palette = "Set1") +
   scale_x_continuous("Month", 1:12, labels = month.abb) +
   scale_y_continuous("# of events", limits = c(0, NA)) +
   facet_wrap(~cPC, ncol = 1)
```


```{r}
oni[events, on = "time"] %>% 
   .[complete.cases(.)]  %>%
   {
         chis <<- .[, broom::tidy(chisq.test(table(enso.event, event))), by = cPC] 
         .
   }  %>% 
   .[, .N, by = .(enso.event, event, cPC)] %>% 
   ggplot(aes(enso.event, event)) +
   geom_raster(aes(fill= N)) +
   geom_label(aes(label = N)) +
   scale_y_discrete("PC event") + 
   scale_x_discrete("ENSO event") +
   facet_wrap(~cPC)
```

```{r}
knitr::kable(chis)
```


# Relación con QS3

```{r calc-qs3eof}
qs3 <- ncep.f(level = 200, lat = -65:-40) %>% 
   .[, FitWave(gh, 3), by = .(time, lat)] %>% 
   .[, .(amplitude = mean(amplitude),
         phase = .SD[amplitude == max(amplitude)]$phase), 
     by = .(time)]

qs3 <- qs.eof$left %>% 
   .[qs.eof$sdev[, .(PC, sd)], on = "PC"] %>% 
   .[as.numeric(PC) %in% 1:4] %>% 
   .[, gh := gh*sd] %>% 
   # cut.eof(1:4) %>% 
   dcast(time ~ PC, value.var = "gh") %>% 
   .[qs3, on = "time"]

compl <- ncep.f(level = 200, lat = -65:-40) %>% 
   .[, fft2(gh, 3), by = .(time, lat)] %>% 
   .[, .(R = mean(R), I = mean(I)), by = time]

compl <- qs.eof$left %>% 
   .[qs.eof$sdev[, .(PC, sd)], on = "PC"] %>% 
   .[, gh := gh*sd] %>% 
   .[as.numeric(PC) %in% 1:4] %>% 
   dcast(time ~ PC, value.var = "gh") %>% 
   .[compl, on = "time"]
```


```{r pc12-ampl, fig.cap = "Relationship between wave 3 amplitude and the first 2 principal components."}
qs3[, .(time, PC1, PC2, amplitude)] %>% 
   melt(id.vars = c("time", "amplitude"), variable.name = "PC") %>% 
   ggplot(aes(amplitude, value)) +
   geom_point(size = 0.5, alpha = 0.5) +
   geom_smooth() + 
   scale_color_brewer(palette = "Set1") +
   scale_x_continuous("Wave 3 amplitude") + 
   scale_y_continuous("PC Value") +
   facet_wrap(~PC, ncol = 1)
```

::: notes

La relación entre los PCs y la amplitud no es evidente así por separado. 

:::

```{r qs3-eof, fig.cap = "Relationship between 2 first EOFs and amplitude of wave 3. a) Monthly mean wave 3 amplitude as a function of PC1 and PC2; contours represent a smooth surface from data points b) Amplitude of monthly mean wave 3 amplitude as a function of PC1 and PC2 magnitude; line is a loess smooth."}
quantiles <- c(.05, 0.25, 0.5, 0.75, 0.95)
b <- quantile(qs3$amplitude, quantiles)

quantile_breaks <- function(probs = seq(0, 1, 0.25), ...) {
   function(x) {
      quantile(x, probs, ...)
   }
}

model <- nls(amplitude ~ a0 + a*sqrt((PC1 - p1_0)^2 + (PC2 - p2_0)^2), 
             data = qs3) 
cm <- as.list(coef(model)) 

p1 <- ggplot(qs3, aes(PC1, PC2)) +
   geom_cross(size = 0.2) +
   stat_rasa(aes(z = amplitude, color = ..level..),
             fun = SmoothContour,
             geom = "contour2", smooth = 0.07,
             breaks = b) +
   # geom_contour(data = grid, aes(z = amplitude, color = ..level..),
   #              binwidth = 10) +
   geom_point(aes(color = amplitude), size = 0.4) +
   scale_color_viridis_c(breaks = b,
                         limits = range(b),
                         guide = guide_colorstrip_bottom(10, 
                                                         inside = TRUE),
                         oob = scales::squish) +
   scale_x_continuous(expand = c(0, 0)) + 
   scale_y_continuous(expand = c(0, 0)) +
   # coord_cartesian(ylim = c(-2000, 2000),
   # xlim = c(-2000, 2000)) +
   coord_equal() +
   labs(tag = "a")  

p2 <- ggplot(qs3, aes(sqrt(PC1^2 + PC2^2), amplitude)) +
   geom_point(alpha = 0.5, size = 0.5) +
   # geom_abline() +
   geom_smooth(method = "loess", formula = y ~ x,
               color = "black", se = T) +
   # geom_line(data = grid) +
   scale_x_continuous(expression(sqrt(PC1^2 + PC2^2))) +
   scale_y_continuous("Wave 3 amplitude") +
   labs(tag = "b") 
p1 + p2
set_last_plot(p1 + p2)
```

::: notes

La relación conjunta entre las variables sí es muy fuerte. Claramente la amplitud de la onda 3 está asociada a la magnitud del vector (PC1, PC2)

:::

```{r phase-PCs, fig.cap = "Phase of zonal wave 3 and value of PCs 1 to 2. The line is a loess smooth."}
smoothcircular <- function(data, range = c(0, 360), span = 0.75, n = 50) {
   data <- rbindlist(list(left = data,
                          mid = data,
                          right = data), 
                     idcol = "side")
   
   setDT(data)
   data[side == "right", x := x + diff(range)]
   data[side == "left", x := x - diff(range)]
   
   x <- seq(min(range), max(range), length.out = n)
   if (max(x) != max(range)) x[n+1] <- max(range)
   
   grid <- data.frame(x = x)
   grid$y <- predict(loess(y ~ x, data = data, span = span), grid)
   grid
}

qs3[, .(time, PC1, PC2, phase, A = Mag(PC2, PC1))] %>% 
   melt(id.vars = c("time", "phase", "A"), variable.name = "PC") %>% 
   ggplot(aes(phase*180/pi, value, color = PC)) +
   geom_point(size = 0.5, alpha = 0.5) +
   # geom_smooth(span = 0.15*3) +
   stat_rasa(fun = smoothcircular, geom = "line", span = 0.15,
             range = c(0, 360/3)) +
   scale_color_brewer(palette = "Set1") +
   scale_x_continuous("Phase", 
                      breaks = seq(0, 360/3, by = 15),
                      limits = c(0, 360/3), 
                      expand = c(0, 0)) +
   scale_y_continuous("PC Value") #+ coord_polar() #+  facet_wrap(~PC)
```

::: notes

PC1 y PC2 también tienen información sobre la fase de la onda 3.

:::


```{r phase-angle, fig.cap = "Relationship between wave 3 phase and angle between PC2 and PC1. Points to the left of the left dotted line are duplicates from the right of the right dotted line. Dashed line is the x = y line and solid line is a linear fit based on the Theil-Sen estimator."}
qs3 %>% copy() %>% 
   .[, pc.angle := ConvertLongitude(atan2(PC2,
                                          PC1)*180/pi)] %>% 
   .[, phase := phase*180*3/pi] %>% 
   rbind(.,., idcol = TRUE) %>%
   .[.id == 1, pc.angle := pc.angle-360] %>%
   .[pc.angle %between% c(-10, 360)] %>%
   .[, dupl := .id == 2] %>% 
   ggplot(aes(pc.angle, phase)) + 
   geom_point(size = 0.4) +
   geom_abline(linetype = 2) +
   geom_smooth(method = "theilsen", color = "black") +
   geom_vline(xintercept = c(0, 350), linetype = 3) +
   scale_y_continuous("Phase x 3 (degrees)") +
   scale_x_continuous(expression(tan^{-1}~(PC2/PC1))) +
   coord_equal()
```

::: notes

The first two principal modes are closely related to the zonal wave 3. In fact, it can be shown that they define a rotation of the complex plane defined by the real an imaginary part of the wave3 fourier transform. 


:::
```{r angle-table}
a1 <- coef(lm(PC1 ~ R + I -1, data = compl))
a2 <- coef(lm(PC2 ~ R + I -1, data = compl))

a1 <- coef(lm(R ~ PC1 + PC2 -1, data = compl))
a2 <- coef(lm(I ~ PC1 + PC2 -1, data = compl))

angles <- as.data.table(rbind(a1, a2))
angles[, PC := c("PC1", "PC2")]
angles[, PC := c("R", "I")]
setcolorder(angles, c(3, 1:2))

knitr::kable(angles)
```

```{r vectors, fig.cap = "Angle between PC1 and PC2 and Real and Imaginary part of wave 3"}
ggplot(angles, aes(PC1, PC2)) +
   geom_segment(aes(x = 0, y = 0, xend = PC1, yend = PC2), 
                arrow = grid::arrow()) +
   coord_equal() 
```

