---
title: "paper de la tesis"
author: "Elio Campitelli"
output: 
   pdf_document: default
   keep_tex: yes
editor_options: 
  chunk_output_type: console
header-includes:
   - \usepackage{subfig}
   - \newcommand*\mean[1]{\overline{#1}}
---

```{r setup, include=FALSE}
# Notification
start.time <- unclass(Sys.time())
min.time <- 10
knit_doc <- knitr::knit_hooks$get("document")
knitr::knit_hooks$set(document = function(x) {
   took <- unclass(Sys.time()) - start.time
   if (unclass(Sys.time()) - start.time >= min.time) {
      notify("Done knitting!", 
             paste0("Took ", round(took), " seconds"),
             time = 5)
   }  
   knit_doc(x)
})

name <- tools::file_path_sans_ext(knitr::current_input())

knitr::opts_chunk$set(echo = FALSE,
                      cache = TRUE, cache.lazy = TRUE,
                      cache.path = paste0("cache/", name, "/"),
                      fig.path = paste0("fig/", name, "/"),
                      warning = FALSE,
                      message = FALSE,
                      fig.align = "center",
                      out.extra = ""
                      # dev = "cairo_pdf"
)

library(data.table)
library(ggplot2)
library(dplyr)
library(metR) 
library(circular)
library(hrbrthemes)
library(extrafont)
library(patchwork)
library(officer)
library(here)
library(ggperiodic)

here <- here::here

knit_error <-  knitr::knit_hooks$get("error")
knitr::knit_hooks$set(error = function(x, options) {
   notify("Error knitting!", time = 5)
   knit_error(x, options)
})


source(here("scripts", "helperfun.R"))
# Plot thingys

data.world <- BuildMap(res = 1, smooth = 1)
map.world <- geom_map2(data.world)
map.SH <- geom_map2(data.world[lat %b% c(-90, 20)], color = "gray20")


pres <- ReadNetCDF(here("DATA", "srfp.mon.mean.nc"))
pres.mean <- pres[, .(pres = median(pres)), by = lat]
pres.mean <- rbind(data.table(lat = 0.0, pres = Inf), 
                   pres.mean, 
                   data.table(lat = -90.0, pres = Inf))
surface <- geom_polygon(data = pres.mean, aes(y = pres), fill = "white", 
                        alpha = 1, color = "gray30", size = 0.5)
pres <- pres[, .(pres = mean(pres)), by = .(lon, lat)]

# From https://github.com/hrbrmstr/hrbrthemes/issues/18
d <- read.csv(extrafont:::fonttable_file(), stringsAsFactors = FALSE)
d[grepl("Light", d$FontName),]$FamilyName <- font_rc_light
write.csv(d,extrafont:::fonttable_file(), row.names = FALSE)
extrafont::loadfonts()

theme_elio <- hrbrthemes::theme_ipsum_rc(base_size = 10) +
   # theme_minimal(base_size = 11) +
   theme(
      legend.position = "bottom", legend.box = "vertical",
      panel.spacing.y = unit(5, "mm"),
      panel.spacing.x = unit(5, "mm"),
      legend.spacing = unit(2, "mm"),
      plot.margin = grid::unit(rep(3, 4), "mm"),
      legend.title = element_blank(),
      legend.box.spacing = unit(3, "mm"),
      legend.margin = margin(t = -5),
      panel.grid = element_line(color = "gray50", size = 0.2, linetype = 3),
      panel.ontop = TRUE)
theme_set(theme_elio)
update_geom_defaults(metR:::GeomTextContour, list(family = font_rc))
hrbrthemes::update_geom_font_defaults()
update_geom_defaults("contour2", list(color = "black"))
update_stat_defaults("contour2", aes(linetype = factor(-sign(..level..))))

options(ggplot2.continuous.fill = "viridis", 
        ggplot2.continuous.color = "viridis")

coord_quickmap <- function(..., ylim = c(-90, -15)) {
   ggplot2::coord_quickmap(ylim = ylim, ...) 
} 

geom_contour_ <- function(..., gap = 0, rotate = FALSE) {
   if (gap != 0) {
      list(geom_contour2(..., gap = gap),
           geom_text_contour(..., rotate = rotate))
   } else {
      geom_contour2(...)
   }
}

# For vertical cross-sections
coord_latlev <- function(ratio = 20, ...) coord_fixed(ratio = ratio, ...)
coord_lonlev <- function(ratio = 20*4, ...) coord_fixed(ratio = ratio, ...)

lev.breaks <- c(1000, 500, 200, 100, 50, 10)

season <- function(...) {
   metR::season(..., lang = "en")
}


knitr::knit_hooks$set(
    rotate = function(before, options, envir) {
        if (before) {
            return("\\begin{landscape}")
        } else{
            return("\\end{landscape}")
        }
    }
)

```

```{r read-ncep}
subset.time <- lubridate::as_datetime(c("1979-12-01", "2017-12-01"))
ncep.f <- memoise::memoise(function(lat = -90:40, 
                                    lon = 0:360,
                                    time = lubridate::as_datetime(c("1979-12-01",
                                                                    "2017-12-01")),
                                    level = 10:1000,
                                    vars = "gh"){
   subset <- list(lat = lat, lon = lon, level = level,
                  time = time)
   n <- ReadNetCDF(here("DATA", "NCEP Reanalysis", "hgt.mon.mean.nc"), vars = c(gh = "hgt"),
                   subset = subset) %>% 
      setnames(., c("level"), c("lev"))
   if ("u" %in% vars) {
      n[, u := ReadNetCDF(here("DATA", "NCEP Reanalysis", "uwnd.mon.mean.nc"), out = "vector",
                          subset = subset)[[1]]] 
   }
   if ("v" %in% vars) {
      n[, v := ReadNetCDF(here("DATA", "NCEP Reanalysis", "vwnd.mon.mean.nc"), out = "vector",
                          subset = subset)[[1]]]
   }
   n[, time := as.Date(time[1]), by = time]
   return(n)
}, cache = memoise::cache_filesystem(".rcache"))
ncep <- ncep.f()
```


```{r calc-fourier1}
rect.annotation <- data.frame(latmin = -65, latmax = -40,
                              levmin = 100, levmax = 700, 
                              k = 3)
gh.season <- ncep[, .(gh = mean(gh)), by = .(lat, lon, lev, season(time))] 
ncep.qs <- ncep[lat != -90, FitWave(gh, 1:3), by = .(lat, lev, time)] %>%
   .[, phi.s := mean.phase(amplitude, phase, k), by = .(k, month(time), lat, lev)] %>% 
   .[, c(mean.wave(amplitude, phase, k)[c("amplitude", "phase")],
         amoma = 2/pi*asin(stationarity.wave2(as.wave(amplitude, phase, k,
                                                      phi.s)))),
     by = .(lat, lev, season(time), k)] 
```

# Introduction

Planetary waves with wave numbers 1 and 3 (QS1 and QS3) are the main zonal asymmetries of the mean circulation of the southern hemisphere (SH) [@Loon1972; @Trenberth1980a]. These zonal waves tend to be quasi-stationary but exhibit significant temporal variability in amplitude and phase [@Loon1972]. In particular, it has been documented that QS3 have an equivalent barotropic structure with  variability in daily, weekly [@Kidson1988], seasonal [@Mo1985] and longer scales [@Karoly1989]. @Mo1985 showed that there exist preferential locations for QS3 action centers. However, due to the age of many of these works, their climatologies have inherent limitations from the small amount of data available in the SH prior to the satellite era.

@Trenberth1985 noted a recurrence in the occurrence of simultaneous blocking anticyclones in different regions of the SH (southern South America, southern New Zealand and central Indian Ocean) favored by the establishment of a QS3 pattern, although they left open the possibility that this is a localized wave train rather than a true zonal wave. @Renwick2005, using clustering techiques, suggestedd that wave 3 blocking was indeed a common occurrence, but noted that this conclusion was sensitive to some data processing decisions. From that time until now, different studies have focused on understanding the influence of blocking on South American climate and their relationship to QS3 [e.g. @Rao2004] or their effect on sea ice [e.g. @Raphael2007] but very little attention has been paid to the study of the causes that lead to the establishment of this pattern in the first place. @Quintanar1995 conducted sensitivity experiments trying to identify the important factors in the maintenance of QS1. They found that the thermal-orographic conditions over Antarctica were not sufficient to explain the QS1 of subpolar latitudes, so they concluded that remote forcers should play an important role. @Wang2013 found that the destruction and recovery of the ozone layer is associated with an increase and decrease in planetary wave activity respectively, but their analysis does not distinguish in the activity of different wave numbers. On the other hand, @Hobbs2010 questioned the usefulness of analyzing SH circulation in zonally symmetric planetary waves and proposed that geopotential zonal anomalies are best characterized by a pair of anticyclones in southern New Zealand and southern South America. 

@Cai1999 evaluated the ability of the CSIRO model to represent QS3 and found that it provided an adequate representation. @Raphael1998 examined QS3 in the NCAR CCM model (versions 1 and 3) and found differences in how each version simulated it, the results being sensitive to the degree of model representation of both sea-atmosphere interaction and sea-ice conditions in polar areas. This is an important result because although QS3 can arrise by the internal dynamics of the atmosphere [e.g. @Simpson2013], there is evidence that surface conditions can influence its activity. Also, @Raphael2003 found important interannual variations experienced by QS3 between 1958 and 1996, apparently related to variations in ENSO frequency, which also influenced SAM asymmetries [@Fogt2012].



## Description and literature


## ZW3 != QS3 

XX Needs redaction:
* Zonal Waves (ZW) and Quasistationary Waves (QS) are related but conceptually distinct.

* ZW is a property of a global scalar field (geopotential in this present case) while QS are a property of *a set* of global scalar fields. 

* QS for a set of fields can be defined as the ZW of the mean field. In that sense, measuring the "activity" of QS waves at a monthly scale is impossible. What *can* be measured is 
  * ZW activity (and asume that is a good proxy for QS activity)
  * monthly field similarity to the QS field --this leads to non-trivial questions about *how* to measure similarity. 

XX


# Methods 

## Data source


## Fourier?

## Steadiness 

Since planetary waves are charecterised by amplitude and phase, they can be treated similarly to other vector quantities like wind. Following this analogy, we measure the degree of stationarity of quasistationary waves by calculating steadiness. That is

$$
\begin{aligned}
k &=  \frac{ |\mathbf{\mean{V}}|}{ \mean{ |\mathbf{V} | } } \\
S &= \frac{2}{\pi}\sin^{-1} \left (  k \right )
\end{aligned}
$$

Where $k$ is the *constancy* and is defined as the ratio between the magnitude of the mean wave ( $|\mathbf{\mean{V}}|$  ) and the mean magnitude of waves $\mean{ |\mathbf{V} | }$. Steadiness ( $S$ ) transforms that value to linearize the $k$ with respect to the phase range [@Singer1967]. A value of 1 represents a completely stationary wave while a value of 0 represents a completely random wave. 

XXX Probably needs better notation.

In practice, `k` can be computed more efficiently by 

$$
k = \frac{\sum_{i=1}^{N}\mathrm{A}_i\cos\left (\mean{\phi} - \phi_i \right )}{\sum_{i=1}^{N}A_i}
$$

Where $A_i$, $\phi_i$ are the amplitude and phase of the ith wave and $\mean{\phi}$ is the phase of the mean wave (stationary phase). That is, $k$ is a weighted mean of the cosine of the difference between phases and the stationary phase. Further, considering that $\cos\left (\mean{\phi} - \phi_i \right )$ is the correlation between two waves of different phase but same frequency makes the interpretation of $k$ clearer. It can be seen that $k$ (and by extension $S$) is a measure of how similar are individual wave 3 fields to the climatological wave 3 field. 

Note that in order to filter out the effect of the monthly seasonal cycle, here $S$ is computed using $\mean{\phi}$ for each month even though the sum is performed for each trimestre. 


XXX Is this too much?

## Wave Action

XXX Needs some definition and methodloty or just a citation to @Vera2004 further along?


# Results

## QS3 climatology

```{r ampl-ncep, fig.cap = "Fourier amplitude of the mean zonal wave 3 (contours) XXX"}
qs.lab <- AddPreffix("QS")

binwidth.ampl <- 10
binwidth.stat <- 0.1
breaks <- 2^seq(0, log2(650), by = 1)
ggplot(ncep.qs[lat <= 0], aes(lat, lev)) +
   geom_contour_fill(aes(z = amplitude), breaks = breaks) + 
   # geom_text_contour(aes(z = amplitude), color = "black", 
   #                   binwidth = binwidth.ampl,
   #                   stroke.color = "#2B828C", stroke = 0.15) +
   surface +
   geom_index.region(rect.annotation) +
   scale_fill_viridis_c(name = "Amplitud (escala logarítmica)", 
                        trans = "log2",
                        breaks = breaks,
                        labels = round(breaks, 1),
                        guide = guide_colorstrip_bottom(),
                        option = "D") +
   scale_y_level(breaks = lev.breaks) +
   scale_x_latitude(name = "latitude", trans = "reverse", ticks = 15) +
   coord_latlev() +
   facet_grid(k ~ season, labeller = labeller(k = qs.lab)) 
```


```{r calc-index}
lats.index <- c(-65, -40)
levs.index <- c(100, 700)

index <- ncep[, qs.index(gh, lat, lev, k = 3), by = time]
```

Figure \ref{fig:ampl-ncep} shows the amplitude of the mean wave for QS1 to 3 for each trimestre. QS1 dominates in almost all levels and latitudes during all seasons but its amplitude is strongest in the austral spring and at higher levels of the atmosphere. QS2, on the other hand... ?

The amplitude of QS3 achieves it's maximum around 200hPa and 50°S during all seasons but with important changes in it's vertical extent. During austral winter and autum, wave 3 amplitude extends up to the higest model levels while it remains restricted to the troposphere and lower stratosphere during spring and summer. 

The tropical upper troposphere stands out as an area of high QS amplitude for the three wavenumbers and almost during all seasons. This is porobably related to the stationary response to the SAMS, specifically the Bolivian High and the xx Low (name??). This can be confirmed by a similar analysis of wave 3 amplitude but using wavelets, which shows that that the area of high QS3 amplitude is located over South America (not shown).  (xx not shown? I don't know if just tease an innovative result among the old methodology.)


```{r calc-gh3-season}
gh.season <- gh.season[, qs3 := FilterWave(gh, 3), by = .(lat, lev, season)]
```

```{r gh3-season, fig.cap = "Wave 3 component of the geopotential field of each season", fig.height=3, fig.subcap=c("At 200hPa", "between 65°S and 35°S"), fig.show="h", fig.ncol = 1}
binwidth <- 7

ggplot(gh.season[lev == 200], aes(lon, lat)) +
   geom_contour_fill(aes(z = qs3), breaks = AnchorBreaks(0, binwidth, 0), 
                     xwrap = c(0, 360)) +
   geom_contour2(aes(z = qs3), breaks = AnchorBreaks(0, binwidth, 0),
                 xwrap = c(0, 360)) +
   map.SH +
   scale_s_map(ylim = c(-90, 20)) +
   scale_fill_divergent(breaks = AnchorBreaks(0, binwidth, 0), 
                        name = "QS3",
                        guide = guide_colorstrip_bottom()) +
   facet_wrap(~season, ncol = 2) +
   coord_quickmap()

cutlats <- c(-65, -35)
binwidth <- 7
ggplot(gh.season[lat %between% cutlats, .(qs3 = mean(qs3)), 
                 by = .(lon, lev, season)], 
       aes(lon, lev)) +
   geom_contour_fill(aes(z = qs3), breaks = AnchorBreaks(0, binwidth, 0)) +
   geom_contour3(aes(z = qs3), breaks = AnchorBreaks(0, binwidth, 0)) +
   scale_y_level(breaks = lev.breaks) +
   scale_x_longitude(name = "lon") +
   scale_fill_divergent(breaks = AnchorBreaks(0, binwidth, 0), 
                        name = "QS3",
                        guide = guide_colorstrip_bottom()) +
   facet_wrap(~season, ncol = 2) 
```

The reconstructed QS3 field at 200hPa shows the expected maximum at 50°S and the anual cycle in its phase already observed in previous studies. It also shows an easterly inclination of the systems evident in all seasons except winter, suggesting that the QS3 might be related to southerly transport of momentum. The vertical strucure of the QS3 shows that the expected seasonal cycle in vertical extent is followed by a slight inclination of the disturances. 

Taken together, this suggests a mix of baroclinic and barotropic processes related to the QS3. The former more important during summer and the the latter dominating in winter.


```{r ampl-stat, fig.cap = "Stat." }
binwidth.ampl <- 10
binwidth.stat <- 0.1
breaks <- 2^seq(0, log2(650), by = 1)
ggplot(ncep.qs[lat <= 0 & k == 3], aes(lat, lev)) +
   geom_contour_fill(aes(z = amoma), breaks = MakeBreaks(0.1)) +
   # geom_contour_tanaka(aes(z = amoma*10), breaks = MakeBreaks(1)) +
   # geom_contour2(aes(z = amoma), breaks = 0.3) +
   geom_contour2(aes(z = amplitude),
                 size = 0.1) +
   # geom_text_contour(aes(z = amoma*10), breaks = c(7, 9), skip = 0, 
   #                   rotate = FALSE, stroke = 0.1, stroke.color = "#2B828C") +
   surface +
   # geom_index.region(rect.annotation) +
   scale_fill_viridis_c(name = "amoma", 
                        # trans = "log2",
                        breaks = MakeBreaks(0.1),
                        limits = c(0, 1),
                        oob = scales::squish,
                        # labels = scales::percent,
                        guide = guide_colorstrip_bottom(),
                        option = "D") +
   scale_y_level(breaks = lev.breaks) +
   scale_x_latitude(name = "latitude", trans = "reverse", ticks = 15) +
   coord_latlev() +
   theme(panel.grid.minor.y = element_blank()) +
   facet_wrap(~ season, ncol = 2)
```

Figure \ref{fig:ampl-stat} show QS3 steadiness for each trimestre of the year. The highest steadiness can be seen in tropical upper troposphere north of 30°S, specially during the summer.

```{r}
ncep[lat <= 0 & month(time) == 11, FitWave(gh, 3), by = .(time, lat, lev)] %>% 
   .[, stationarity.wave2(as.wave(amplitude, phase, k)), by = .(lat, lev)] %>% 
   .[, V2 := 2/pi*asin(V1)] -> nov
```


With the exception of spring, areas of high amplitude (shown in black contours in Figure \ref{fig:ampl-stat}) are colocated with areas of high steadiness --although this metric is maximized at lower levels. While the mean wave amplitude in spring is high (not shown), steadiness is low in the midlatitudes and specially in the stratosphere. This is due primarily by the behaviour during november, when steadiness is so low (reaching ~`r round(nov[lat == -60 & lev == 500]$V2, 2)` at 60°S, 500hPa) that wave 3 arguably cannot be considered stationary in the midlatitudes. 

As described before, monthly ZW activity can be used as a proxy for QS activity at a monthly scale. Following this an index for the activity of ZW3 was constructed by taking the average amplitude and phase of the wave 3 component in a box with limits between 45°S and 60°S and between 100hPa and 700hPa (shown as a dotted box un Figure \ref{fig:ampl-ncep}). The seasonal cycle of the amplitude componet is shown in Figure \ref{fig:boxplot-index} along with the seasonal cycle of the amplitude of the QS3 wave in the same box.  

```{r}
ncep[lat %between% c(-60, -45) & lev %between% c(100, 700), 
     mean(gh), by = .(lon, lat, lev, month(time))] %>% 
   .[, FitWave(V1, 3), by = .(lat, lev, month)] %>% 
   .[, .(QS3 = mean(amplitude)), by = .(month)] -> qs3
```


```{r boxplot-index, fig.cap = "Seasonal cicle of QS3 index."}
set.seed(42)
ggplot(index, aes(factor(month(time)), amplitude)) +
   # geom_violin() +
   geom_boxplot(outlier.alpha = 0,  aes(fill = season(time))) +
   geom_line(data = qs3, aes(x = month, y = QS3)) +
   geom_sina(alpha = 0.5, size = 0.7) +
   scale_x_discrete(name = "Month", labels = month.abb) +
   scale_y_continuous(name = "Amplitude") +
   scale_fill_brewer(palette = "OrRd", guide = "none") +
   theme(panel.grid.major.x = element_blank())
```

ZW3 activiy exhibits a clear annual cycle with higher amplitude during the austral winter months and lower amplitude during the summer months. This annual cycle is consistent with previous work [@Loon1972; @Karoly1985; @Raphael2004]. There is also an annual cycle in variability with winter months showing a high degree of variability compared to summer. 

There is no evidence of persistence, periodicities or xxlong time tendencies (not shown).

A comparison with the QS3 cycle (black line in Figure \ref{fig:boxplot-index}) echoes what is seen in Figure \ref{fig:ampl-stat}. Winter months have a low level of steadiness that makes the ZW and QS properties diverge. As suggested previously, the low amplitude of QS3 wave in spring is not due to a reduction in the amplitude of the planetary wave 3 but to its reduced steadiness. This divergence should guide the interpretation of results based on the index. 

```{r fase-boxplot-arg}
lat.lims <- c(-60, -22)
lon.lims <- c(240, 360)

map.data.arg <- BuildMap(countries = T) %>% 
   .[lat %between% c(lat.lims + c(0, 10)) & long %between% lon.lims]


map <- function(data) {
   geom_polygon(data = data, aes(long, lat, group = group),
                fill = "gray70", color = "white", size = 0.1,
                inherit.aes = F)
}

nudge <- 0.54
size <- 2

index <- index[, phase.c := circular(phase*3, modulo = "2pi")]

index_points <- index %>%
   group_by(month = month(time)) %>%
   filter(greater(amplitude, 20)) %>% 
   mutate(phase = phase*180/pi) %>% 
   ggperiodic::periodic(phase = c(0, 120)) %>% 
   ggperiodic::wrap(phase = c(0, 360))

arg <- index %>% 
   .[, .(phase = mean.wave(amplitude, phase, 3)$phase,
         phase.sd = sd.circular(phase.c)/3),
     by = .(month(time))] %>% 
   ggplot(aes(y = (-month + 6)*2.7 - 45)) +
   map(map.data.arg) +
   geom_point(data = index_points,
              aes(x = phase),
              alpha = 0.8, size = size, color = "red") +
   geom_point(data = index_points,
              aes(x = phase + 60),
              alpha = 0.8, size = size, color = "blue",
              shape = 5,
              position = position_nudge(y = nudge)) +
   geom_point(aes(phase*180/pi + 240), shape = 0) +
   geom_errorbarh(aes(xmin = (phase*180/pi + 240 - phase.sd*180/pi), 
                      xmax = (phase*180/pi + 240 + phase.sd*180/pi))) +
   geom_vline(xintercept = c(240, 360), linetype = 3) +
   scale_x_longitude(ticks = 20, name = "Longitude") +
   scale_y_continuous(breaks = (-(1:12) + 6)*2.6 - 45, 
                      minor_breaks = NULL,
                      labels = month.abb,
                      name = "", expand = c(0, 0)) +
   coord_quickmap(xlim = lon.lims, ylim = c(-65, -25)) 
```

```{r fase-boxplot-nz}
lon.lims.nz <- c(120, 240) - 20

map.data.nz <- BuildMap(countries = T) %>% 
   .[lat %between% (lat.lims + c(0, 10)) & 
        long %between% lon.lims.nz] %>% 
   .[, lat := lat]

map.nz <- geom_polygon(data = map.data.nz, aes(long, lat, group = group),
                       fill = "gray90", color = "black", size = 0.1,
                       inherit.aes = F)

nz <- index %>% 
   .[, .(phase = mean.wave(amplitude, phase, 3)$phase*180/pi + 120,
         phase.sd = sd.circular(phase.c)/3),
     by = .(month(time))] %>% 
   ggplot(aes(y = (-month + 6)*2.7 - 45)) +
   map(map.data.nz) +
   geom_point(data = index_points,
              aes(x = phase),
              alpha = 0.8, size = size, color = "red") +
   geom_point(data = index_points,
              aes(x = phase + 60),
              alpha = 0.8, size = size, color = "blue",
              shape = 5,
              position = position_nudge(y = nudge)) +
   geom_point(aes(phase), size = size + 1, shape = 0) +
   geom_errorbarh(aes(xmin = (phase - phase.sd*180/pi), 
                      xmax = (phase + phase.sd*180/pi))) +
   geom_vline(xintercept = lon.lims.nz, linetype = 3) +
   scale_x_longitude(ticks = 20, name = "Longitude", position = "top") +
   scale_y_continuous(breaks = (-(1:12) + 6)*2.6 - 45, 
                      minor_breaks = NULL,
                      labels = month.abb, 
                      name = "", expand = c(0, 0)) +
   coord_quickmap(xlim = lon.lims.nz, ylim = c(-65, -25)) 
```

```{r fase-boxplot, fig.cap = "Phase of the mean wave for each month and $\\pm$ 1 standrad deviation (black dots and error bars). In red and blue, phase of the most intense 20 events for each month, representing the location of the ridge and throuht respectively."}
nz + arg + plot_layout(ncol = 1) + plot_annotation(tag_levels = "a", tag_suffix = ")")
```

Figure \ref{fig:fase-boxplot} shows the mean phase of ZW3. Red circles and blue diamonds represent the position of the geopotential maximum and minimum, respectively. Mean position and $\pm$ 1 standard deviation is shown by black squares and error bars. (XX more about background and x y axis?). The importance of the phase variability in ZW3 can be apreciated in the context of the position of throughs and ridges in relationship to the continents. Ridges can be to the west or to the east of Oceania and South America even on the same month of the year. These very different forced circulations have oposite effects (i.e. advection of cold or warm air), which underscores not only the importance of taking location into account, but also the possiblity that this variability could mask those impacts at a climatological level. Also of note is the non-trivial cases of throughs located in areas climatologically high pressure areas. This is specially evident in november. 

Ignoring for now the issues of representation, ZW3 index can be regressed with various atmospheric variables to infer the typical structure of the circulation related to high ZW3 activity.

```{r read-psi}
psi.file <- here("DATA", "NCEP Reanalysis", "psi.mon.mean.nc")
levs <-  ReadNetCDF(psi.file, out = "vars")$dimensions$level
levs <- levs[levs %~% 0.21]
stream <- ReadNetCDF(psi.file, subset = list(lat = -90:40,
                                             level = levs,
                                             time = subset.time)) %>% 
   # setnames(c("level", "time"), c("lev", "date")) %>% 
   .[, psi.z := Anomaly(psi), by = .(lat, time)]
stream[, time := as.Date(time[1]), by = time]
```

```{r read-sst}
sst <- ReadNetCDF(here("DATA", "sst.mnmean.nc"), 
                  subset = list(time = subset.time, 
                                lat = -90:20), vars = "sst")
sst[, time := as.Date(time[1]), by = time]
```


```{r calc-regrs}
index_sd <- copy(index) %>% 
   .[, amplitude := amplitude/sd(amplitude)]


sst.reg <- sst[index_sd, on = "time"] %>% 
   # .[complete.cases(.)] %>%
   .[, FitLm(sst, amplitude, phase), by = .(lon, lat, month(time))] %>% 
   .[term == "amplitude"] 

sst.reg[, estimate_impute := Impute2D(estimate ~ lon + lat), by = month]

stream.reg <- stream[, .(lon, lat, time, psi.z)] %>% 
   .[index_sd, on = "time"] %>% 
   .[, FitLm(psi.z, amplitude, phase = phase, 
             se = TRUE), 
     by = .(lat, lon, month(time))] %>% 
   .[term == "amplitude"] %>% 
   .[, psi.z := estimate] %>% 
   .[, c("f.x", "f.y") := WaveFlux(.SD), by = .(month)] %>% 
   .[complete.cases(.)]

gh.reg <- ncep[lev == 200][index_sd, on = "time"] %>% 
   .[, FitLm(gh, amplitude, phase), 
     by = .(lat, lon, month(time))] %>% 
   .[term == "amplitude"]
```


```{r regressions, fig.cap = "Regression between scaled ZW3 index and SST (shaded) and geopotential height (contours), and wave activity flux derived from regression between ZW3 index and streamfunction at 0.2101 sigma (streamlines).", fig.height=(297 - 10)/24.5, fig.width=(210-10)/24.5}
months <- 1:12 # c(1, 3, 4, 7, 8, 12)
binwidth <- 10
stream.g <- stream.reg[lat <= 15 & month %in% months] 
gh.g <- gh.reg[month %in% months] #%>% 
   # .[, estimate := Anomaly(estimate), by = .(month, lat)]

gh.g <- periodic(gh.g, lon = c(0, 360))
stream.g <- periodic(stream.g, lon = c(0, 360))
sst.reg <- periodic(sst.reg, lon = c(0, 360))

ggplot(gh.g, aes(lon, lat)) +
   geom_contour_fill(data = sst.reg, aes(z = estimate), na.fill = 0,
                     breaks = AnchorBreaks(0, 0.1, 0))  +
   geom_contour2(data = sst.reg, aes(z = estimate, 
                                     linetype = factor(-sign(stat(level)))),
                  breaks = AnchorBreaks(0, 0.1, 0), na.fill = 0, size = 0.1) +
   stat_subset(data = sst.reg, aes(subset = is.na(estimate)), fill = "white", 
               geom = "raster") +
   geom_contour2(aes(z = estimate, linetype = factor(-sign(stat(level)))),
                 breaks = AnchorBreaks(0, binwidth, 0),
                 size = 0.5) +
   geom_streamline(aes(dx = f.x, dy = f.y), skip = 5, min.L = 3,
                   xwrap = c(0, 360),
                   res = 4, L = 10, arrow.length = 0.2, size = 0.2,
                   data = stream.g) +
   map.world +
   scale_s_map(ylim = c(-90, 15)) +
   scale_fill_divergent(breaks = AnchorBreaks(0, 0.1, 0),
                        na.value = "white",
                        guide = guide_colorstrip_bottom()) +
   coord_quickmap(ylim = c(-90, 15)) +
   facet_wrap(~month, dir = "h", ncol = 2,
              labeller = labeller(month = month.abb)) +
   # facet_wrap_paginate(~month, page = 1, nrow = 1, ncol = 1)
   theme(panel.spacing.y = unit(0, "lines"))
```

Figure \ref{fig:regressions} shows regressions between scaled ZW3 index and SST, and 200hPa geopotential height and wave activity flux derived from regressions between scaled ZW3 index and streamfunction at `r levs` $\sigma$ for each month. All months show evidence of planetary wave 3 activity, but with considerable heterogeneity in its characteristics. In particular, it can be seen that some months (January, May, September and December) are characterised by a wave train emanating from the tropical western Pacific propagating to the southeast and then turning to the northeast. Other months (March, April, July and October) show a much more zonally coherent wave 3 pattern that extends along latitude circles without such a dramatic meridional propagation. 

Patterns of SST variation are similarly mixed. There's a strong signal in the equatorial pacific, but it is not consistent, varying between positive (February, May and September) and negative (January, March, November and December). This variability is masked when performing regressions or compositions for all months together (not shown) and can probably explain the small signal mentioned in @Renwick2005. 

The aforementioned problems of usind ZW3 amplitude as a measure of QS3 activity puts into question the fundamental validity of the regression analysis using this variable. 


# Discussion

ZW amplitude is not a good measure for characterising QS behaviour for several reasons. By ignoring the phase component, it captures waves with one common wavenumber that nevertheless are very different and are exited by different procesess. This problem is exacerbated for periods or months that exibit low levels of steadiness. Moreover, fourier decomposition imposes a sinusoidal form and restriscts the analisys to waves at constant latitude. However, actual waves in the atmosphere can have meridional displacements and different amplitudes within the same latitude circle (Figure??). 

Some alternatives seem promising. 

Applying wavelets in the spatial domain relaxes the sinusoidal form restriction in fourier and allows amplitude to vary with longitude (Figure?). With this information, it becomes possible to distinguish between trully planteary-scale waves and more localized wave paquets. In the atmospherics sciences, wavelets are used extensively in timeseries analisys [ej. @Raphael2004; @Kinnard2011] and image processing [ej. @Desrochers1999], but it's not often used in the spatial domain [ej. @Pinault2016]. It is used in the ecological literautre [ej, @Mi2005].

